---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---


```{r}
library(picante)
```

Community data
```{r}
comm <- read.csv("grassland_data/grassland.community.csv", header = TRUE, row.names = 1)
class(comm)
dim(comm)
rownames(comm)
head(colnames(comm))
comm[1:5, 1:5]
apply(comm, 1, sum)
# Turn percent cover to relative abundance by dividing each value by sample
# total abundance
comm <- decostand(comm, method = "total")
apply(comm, 1, sum)
comm[1:5, 1:5]
```

Trait Data
```{r}
traits <- read.csv("grassland_data/species.traits.csv", header = TRUE, row.names = 1)
head(traits)
pairs(traits)
traits <- log10(traits)
# plot the transformed data
pairs(traits)

```

Metadata
```{r}
metadata <- read.csv("grassland_data/plot.metadata.csv", header = TRUE, row.names = 1)
# take a peek at the data
head(metadata)
```

Phylogeny
If you have a phylogeny in the commonly used Newick or Nexus format it can be imported into R with the read.tree or read.nexus functions.

A phylo object is a special type of list object - it has different elements such as tip labels and edge lengths, and R knows how to summarize and plot a phylo object due to the way it is defined by the ape package.
```{r}
phy <- read.tree("grassland_data/grassland.phylogeny.newick")
class(phy)
phy
names(phy)
phy$tip.label[1:5]
Ntip(phy)
plot(phy, cex = 0.5)
```

Cleaning and matching data sets
The data sets we are using today have already been cleaned up.
```{r}
# check for mismatches/missing species
combined <- match.phylo.comm(phy, comm)
# the resulting object is a list with $phy and $comm elements.  replace our
# original data with the sorted/matched data
phy <- combined$phy
comm <- combined$comm

combined <- match.phylo.data(phy, traits)
# the resulting object is a list with $phy and $data elements.  replace our
# original data with the sorted/matched data
phy <- combined$phy
traits <- combined$data

all.equal(rownames(comm), rownames(metadata))

# they all match - if they didn't we could sort them to the same order sort
# metadata rows to be in the same order as community rows
metadata <- metadata[rownames(comm), ]
```

Community richness and diversity
**Instead of examining the effect of habitat (metadata$habitat) do an analysis of rel.moisture
(metadata$rel.moisture). All the code that examine habitat should be changed to examine
rel.moisture**
```{r}
boxplot(specnumber(comm) ~ metadata$rel.moisture, ylab = "# of species")
metadata$rel.moisture<-as.factor(metadata$rel.moisture) #moisture seems to be in levels so changed that to factor (it was an int)
m1<-lm(specnumber(comm) ~ metadata$rel.moisture)
anova(m1)

plot(specaccum(comm), xlab = "# of samples", ylab = "# of species")

```

Multivariate Community Analysus
```{r}
# calculate Bray-Curtis distance among samples
comm.bc.dist <- vegdist(comm, method = "bray")
# cluster communities using average-linkage algorithm
comm.bc.clust <- hclust(comm.bc.dist, method = "average")
# plot cluster diagram
plot(comm.bc.clust, ylab = "Bray-Curtis dissimilarity")
```

Ordination
```{r}
# The metaMDS function automatically transforms data and checks solution
# robustness
comm.bc.mds <- metaMDS(comm, dist = "bray")
# Assess goodness of ordination fit (stress plot)
stressplot(comm.bc.mds)
# plot site scores as text
ordiplot(comm.bc.mds, display = "sites", type = "text")
# automated plotting of results - tries to eliminate overlapping labels
ordipointlabel(comm.bc.mds)
```
In the chunk above we used the metaMDS


```{r}
# ordination plots are highly customizable set up the plotting area but
# don't plot anything yet
mds.fig <- ordiplot(comm.bc.mds, type = "none")
# plot just the samples, colour by habitat, pch=19 means plot a circle
points(mds.fig, "sites", pch = 19, col = "green", select = metadata$habitat == 
    "Fescue")
points(mds.fig, "sites", pch = 19, col = "blue", select = metadata$habitat == 
    "Mixedgrass")
# add confidence ellipses around habitat types
ordiellipse(comm.bc.mds, metadata$habitat, conf = 0.95, label = TRUE)
# overlay the cluster results we calculated earlier
ordicluster(comm.bc.mds, comm.bc.clust, col = "gray")

# plot Sphaeralcea abundance. cex increases the size of bubbles.
ordisurf(comm.bc.mds, comm[, "Sphaeralcea_coccinea"], bubble = TRUE, main = "Sphaeralcea coccinea abundance", 
    cex = 3)
```

Adding environmental and trait data to ordinations
How are environmental variables correlated with the ordination axes?
```{r}
ordiplot(comm.bc.mds)
# calculate and plot environmental variable correlations with the axes use
# the subset of metadata that are environmental data
plot(envfit(comm.bc.mds, metadata[, 3:6]))
```

Trait evolution
Phylogenetic Signal
```{r}
# one way to do it - apply the Kcalc function to each column of the
# data.frame
apply(traits, 2, Kcalc, phy)

# another way to do it with significance testing we have to convert the tree
# to be dichotomous before calculating P-values
multiPhylosignal(traits, multi2di(phy))


```

Visualizing trait evolution
```{r}
# Plot phylogeny facing upwards. Show node labels but not tip labels. cex
# shrinks labels.
plot(phy, direction = "up", show.tip.label = FALSE, show.node.label = TRUE, 
    cex = 0.7)
# Plot leaf area on the phylogeny. cex argument scales symbol size by trait
# value.
tiplabels(pch = 19, col = "black", cex = 3 * (traits[, "LeafArea"]/max(traits[, 
    "LeafArea"])))
```

Phylogenetic Analysis of trait relationships
```{r}
# GLS of root tissue density as a function of SRL - non-phylogenetic model
root.gls <- gls(RootTissueDens ~ SRL, data = traits)
anova(root.gls)

# Phylogenetic GLS - adds effect of phylogeny to the model
root.pgls <- gls(RootTissueDens ~ SRL, correlation = corBrownian(value = 1, 
    phy), data = traits)
anova(root.pgls)
```

```{r}
# plot relationship
plot(RootTissueDens ~ SRL, data = traits, xlab = "SRL (specific root length)", 
    ylab = "Root tissue density")
# add model fit lines - coef is the model fit coefficients, lwd increases
# line width
abline(coef(root.gls), lwd = 2, col = "black")
abline(coef(root.pgls), lwd = 2, col = "red")
legend("bottomleft", legend = c("GLS fit", "Phylogenetic GLS fit"), lwd = 2, 
    col = c("black", "red"))
```

Phylogenetic and trait diversity
Phylogenetic diversity
```{r}
# Calculate Faith's PD
comm.pd <- pd(comm, phy)
head(comm.pd)
boxplot(comm.pd$PD ~ metadata$rel.moisture, xlab = "Rel. Moisture", ylab = "Faith's PD")
# Test for PD differences among moisture
m2<-lm(comm.pd$PD ~ metadata$rel.moisture)
anova(m2)
summary(m2)
# Compare PD and species richness
plot(comm.pd$PD ~ comm.pd$SR, xlab = "Species richness", ylab = "Faith's PD")
```

MPD, MNTD, SESMPD and SESMNTD
```{r}
# convert phylogenety to a distance matrix
phy.dist <- cophenetic(phy)
# calculate ses.mpd
comm.sesmpd <- ses.mpd(comm, phy.dist, null.model = "richness", abundance.weighted = FALSE, 
    runs = 999)
head(comm.sesmpd)

# calculate ses.mntd
comm.sesmntd <- ses.mntd(comm, phy.dist, null.model = "richness", abundance.weighted = FALSE, 
    runs = 999)
head(comm.sesmntd)

# compare ses.mpd between moisture
plot(comm.sesmpd$mpd.obs.z ~ metadata$rel.moisture, xlab = "Rel. Moisture", ylab = "SES(MPD)")
abline(h = 0, col = "gray")
m4<-lm(comm.sesmpd$mpd.obs.z ~ metadata$rel.moisture)
anova(m4)
summary(m4)
# compare ses.mntd between moisture
plot(comm.sesmntd$mntd.obs.z ~ metadata$rel.moisture, xlab = "Rel. moisture", ylab = "SES(MNTD)")
abline(h = 0, col = "gray")
m5<-lm(comm.sesmntd$mntd.obs.z ~ metadata$habitat)
anova(m5)
summary(m5)
# plot species present in a fescue community
plot(phy, show.tip.label = FALSE, main = "Fescue community fes-K-11")
tiplabels(tip = which(phy$tip.label %in% colnames(comm)[comm["fes-K-11", ] > 
    0]), pch = 19)

# plot species present in a mixedgrass community
plot(phy, show.tip.label = FALSE, main = "Fescue community mix-H-23")
tiplabels(tip = which(phy$tip.label %in% colnames(comm)[comm["mix-H-23", ] > 
    0]), pch = 19)
```

Trait Diversity
```{r}
# calculate trait distance - Euclidean distance among scaled trait values -
# we want the full distance matrix
trait.dist <- as.matrix(dist(scale(traits), method = "euclidean"))
# calculate trait ses.mpd
comm.sesmpd.traits <- ses.mpd(comm, trait.dist, null.model = "richness", abundance.weighted = FALSE, 
    runs = 999)
# compare trait ses.mpd between moisture levels
plot(comm.sesmpd.traits$mpd.obs.z ~ metadata$rel.moisture, xlab = "Habitat", ylab = "Trait SES(MPD)")
abline(h = 0, col = "gray")


```

Phylogenetic beta-diversity
```{r}
# calculate phylogenetic MNTD beta diversity
comm.mntd.dist <- comdistnt(comm, phy.dist, abundance.weighted = TRUE)
# calculate functional trait MNTD beta diversity
comm.mntd.traits.dist <- comdistnt(comm, trait.dist, abundance.weighted = TRUE)
# calculate Mantel correlation for taxonomic Bray-Curtis vs. phylogenetic
# MNTD diversity
mantel(comm.bc.dist, comm.mntd.dist)

# calculate Mantel correlation for taxonomic Bray-Curtis vs. trait MNTD
# diversity
mantel(comm.bc.dist, comm.mntd.traits.dist)
```

Phylogeny/trait-based ordinations
```{r}
# NMDS ordination of phylogenetic distances - use monoMDS since we only have
# among-sample distances
comm.mntd.mds <- monoMDS(comm.mntd.dist)
# set up the plotting area but don't plot anything yet
mds.fig <- ordiplot(comm.mntd.mds, type = "none")
# plot just the samples, colour by habitat, pch=19 means plot a circle
points(mds.fig, "sites", pch = 19, col = "green", select = metadata$habitat == 
    "Fescue")
#points(mds.fig, "rel.moisture", pch = 19, col = "green", select = metadata$rel.moisture == 
   # "Moisture")
points(mds.fig, "sites", pch = 19, col = "blue", select = metadata$habitat == 
    "Mixedgrass")
# add confidence ellipses around habitat types
ordiellipse(comm.mntd.mds, metadata$habitat, conf = 0.95, label = TRUE)


# NMDS ordination of trait distances - use monoMDS since we only have
# among-sample distances
comm.mntd.traits.mds <- monoMDS(comm.mntd.traits.dist)
# set up the plotting area but don't plot anything yet
mds.fig <- ordiplot(comm.mntd.traits.mds, type = "none")
# plot just the samples, colour by habitat, pch=19 means plot a circle
points(mds.fig, "sites", pch = 19, col = "green", select = metadata$habitat == 
    "Fescue")
points(mds.fig, "sites", pch = 19, col = "blue", select = metadata$habitat == 
    "Mixedgrass")
# add confidence ellipses around habitat types
ordiellipse(comm.mntd.traits.mds, metadata$habitat, conf = 0.95, label = TRUE)
```

Testing for multivariate differences among groups
```{r}
# Taxonomic (Bray-Curtis) dissimilarity explained
adonis(comm.bc.dist ~ habitat, data = metadata)
# Phylogenetic dissimilarity explained
adonis(comm.mntd.dist ~ habitat, data = metadata)
# Trait dissimilarity explained
adonis(comm.mntd.traits.dist ~ habitat, data = metadata)
```



