---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r}

testdata<-read.table("RTestData-1.txt",header=TRUE)
attach(testdata)
testdata$time<-as.factor(testdata$time)
```

Normality check & data transformation
(1) Make histograms, run qqnorm and qqline of all bacteria and “deepest”.
(2) Transform variables if non-normal and repeat histogram, qqnorm, qqline on
transformed data. For transformations I suggest trying either sqrt() or log().
(3) Example: lepto=sqrt(lepto

```{r normality check}
hist(strep)
qqnorm(strep)
qqline(strep)
shapiro.test(strep) #not normal

hist(prev)
qqnorm(prev)
qqline(prev)
shapiro.test(prev) #not normal

hist(lepto)
qqnorm(lepto)
qqline(lepto)
shapiro.test(lepto) #not normal

hist(fuso)
qqnorm(fuso)
qqline(fuso)
shapiro.test(fuso) #normal!

hist(veil)
qqnorm(veil)
qqline(veil)
shapiro.test(veil) #not normal

hist(deepest)
qqnorm(deepest)
qqline(deepest)
shapiro.test(deepest) #nice and normal! 

```

```{r transformations and normality check}
#strep
lstrep<-log(strep)
qqnorm(lstrep)
hist(lstrep)
shapiro.test(lstrep) #nice and normal

#prev
lprev<-log(prev+1)
qqnorm(lprev) #error in plot.window(...) : need finite 'ylim' values
hist(lprev)
shapiro.test(lprev) #NA

sprev<-sqrt(prev+1)
qqnorm(lprev) #error in plot.window(...) : need finite 'ylim' values
qqline(lprev)
hist(lprev)
shapiro.test(lprev) #NA

rprev<-rank(prev)
qqnorm(rprev)
hist(rprev)
shapiro.test(rprev)

#lepto
loglepto<-log(lepto)
hist(loglepto)
qqnorm(loglepto)
qqline(loglepto)
shapiro.test(loglepto)

slepto<-sqrt(lepto)
hist(slepto)
qqnorm(slepto)
qqline(slepto)
shapiro.test(slepto) #normal!

```

One-way ANOVA and summary and boxplots:
(1) Lepto by time
(2) Strep by time
```{r Lepto by time}
m1<-lm(slepto~time)
anova(m1) #pvalue 0.02 
summary(m1)
boxplot(slepto,time)
```

```{r Strep by time}
m2<-lm(lstrep~time)
anova(m2) #pvalue 0.3
summary(m2)
boxplot(lstrep,time)

```

Correlation analysis (cor.test)
(1) Strep vs. prev
(2) Strep vs. fuso
(3) Fuso vs. lepto

```{r}
#(1) Strep vs. prev
cor.test(lstrep,prev) #pvalue 0.002
#(2) Strep vs. fuso
cor.test(lstrep,fuso) #pvalue 6.981e-05 #highly correlated
#(3) Fuso vs. lepto
cor.test(fuso,slepto) #pvalue 0.7453

```

Linear regressions
Use the lm command and use plot and abline to produce graphs.
(1) Strep vs. deepest
(2) Fuso vs. deepest
```{r}
#(1) Strep vs. deepest
m3<-lm(lstrep~deepest)
plot(m3)
abline(m3)
#(2) Fuso vs. deepest
m4<-lm(fuso~deepest)
plot(m4)
abline(m4)
```

Produce the following graphs:
(1) ggplot histogram of prev variable.
(2) ggplot scatterplot of strep by deepest, colored by time
(3) ggplot scatterplot of fuso by deepest
(4) Same as 3 but use smoothing algorithm

```{r}
library(ggplot2)
histo<-ggplot(data=testdata, aes(x = prev))+geom_histogram()
histo
```

```{r}
sp<-ggplot(data=testdata,aes(x=deepest,y=lstrep,color = time))+geom_point()
sp
```

```{r}
sp2<-ggplot(data=testdata, aes(x = deepest, y = fuso)) + geom_point()
sp2
```

```{r}
sp3<-ggplot(data=testdata, aes(x = deepest, y = fuso)) + 
  geom_point()+geom_smooth(method = lm)
sp3
```

Part C: Vegan analysis

```{r}
library(vegan)
```
1) Calculate alpha diversity (Shannon) for the dataset.
(2) Rarefy the data and make a plot using rarecurve.
(3) Calculate Bray-Curtis dissimilarity matrix, show the matrix and plot a histogram.
```{r}
newdata=testdata[2:6]
shannon <- diversity(newdata)
shannon
```

```{r}
raremin <- min(rowSums(newdata))
raremin
newdata$strep<-as.integer(newdata$strep+1)
newdata$prev<-as.integer(newdata$prev+1)
newdata$lepto<-as.integer(newdata$lepto+1)
newdata$fuso<-as.integer(newdata$fuso+1)
newdata$veil<-as.integer(newdata$veil+1)
sRare <- rarefy(newdata, raremin) #function accepts only integers (counts)
sRare
rarecurve(newdata, col = "blue")
```

```{r}
bray = vegdist(newdata, "bray")
bray
hist(bray)
```

(1) Make a multidimensional scaling plot using the Bray-Curtis matrix.
(2) Color the plot by status, then time.

```{r}
mds <- metaMDS(comm = newdata, distance = "bray", trace = FALSE, autotransform = FALSE)
plot(mds$points)
MDS_xy <- data.frame(mds$points)

MDS_xy$status <- testdata$status
MDS_xy$time <- testdata$time

#status plot
ggplot(MDS_xy, aes(MDS1, MDS2, color = status)) +
  geom_point() +
  theme_bw()

#time plot
ggplot(MDS_xy, aes(MDS1, MDS2, color = time)) +
  geom_point() +
  theme_bw()
```

